Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)

### 2014-11-18 docs/papers/shallow_performance/shallow_perf.tex
Add detailed discussion of Gnu, Intel and Cray perf evolution

Having show that, in general, we can recover and occasionally improve
upon the performance of the original version of shallow, the next
logical step is to examine the necessary code transformations.  We do
this for the $256^{2}$ case since this fits within cache on both the
Haswell and Ivy Bridge CPUs.  Table~\ref{TABLE_opt_breakdown} shows
detailed timings for this case after each transformation has been
applied to the code. The same data is visualised in
Figure~\ref{FIG_opt_stages_256} where the performance of the best
\psykal version for a given compiler/CPU is given relative to the
performance of the original with the same compiler/CPU. For the Gnu
compiler, the vanilla \psykal version of shallow is almost 20 times
slower than the original. Copying the code for each of the kernels
into the module from which they are called has the most dramatic
effect on the performance of the compiled code. The next most
significant transformation is that of (manually) in-lining the kernel
code into the bodies of the loops from which they are called. This
gets the \psykal version to within 8\% of the performance of the
original.

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Compiler:          &  Gnu    & Intel   & Gnu & Cray & Intel & IBM bgxlf \\
\hline
CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
\hline
Original                             & 1.49 & 1.41 & 1.54 & 1.23 & 1.57  &        \\
Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.67 & 1.96  &        \\
Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.39 & 1.93  & 113.74 \\
Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.42 & 1.93  &  68.28 \\
Loops fused                          & 2.18 & 1.58 & 1.79 & 1.47 & 1.76  &  71.91 \\
In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.46 & 1.75  &  83.36 \\
Fused field-copy                     & 2.18 & 1.85 & 1.99 & 1.30 & 2.07  &  72.62 \\
In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.30 & 1.59  &        \\
Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.25 & 1.55  &        \\
Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.23 & 1.54  &        \\
\hline
\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
\hline
\end{tabular}
\end{table*}

% 0.32/1.41 = 23% and 0.39/1.57 = 25%
Considering now the Intel compiler, Table~\ref{TABLE_opt_breakdown}
shows that even the vanilla \psykal version of shallow is only
23--25\% slower than the original version. The only code
transformation that produced a significant performance improvement for
this case was the fusion of the computational loops. Further
investigation of the slow-down caused by attempting to fuse the
field-copy operation with the preceeding time-smoothing loop revealed
that the compiler was unable to vectorise the loop after it had
in-lined the body of the time-smoothing kernel. Once this in-lining
was done by hand, the compiler was able to vectorise the loop and the
performance of the original version was recovered.

\begin{figure}
\begin{verbatim}
! Time smoothing
DO J=1,N+1
  DO I=1,M+1
    CALL time_smooth_code(i,j,ufld,unew,uold)
!    uold(i,j) = ufld(i,j) + alpha* &
!        (unew(i,j) - 2.*ufld(i,j) + uold(i,j))

    CALL time_smooth_code(i,j,vfld,vnew,vold)
!    vold(i,j) = vfld(i,j) + alpha* &
!        (vnew(i,j) - 2.*vfld(i,j) + vold(i,j))

    CALL time_smooth_code(i,j,pfld,pnew,pold)
!    pold(i,j) = pfld(i,j) + alpha* &
!        (pnew(i,j) - 2.*pfld(i,j) + pold(i,j))
  END DO
END DO

! Copy new fields into current fields
DO J=1,N+1
  DO I=1,M+1
    Ufld(I,J) = UNEW(I,J)
    Vfld(I,J) = VNEW(I,J)
    Pfld(I,J) = PNEW(I,J)
  END DO
END DO
\end{verbatim}
\caption{Example of the coding structure that required manual
  in-lining of the kernel body (as indicated by commented-out lines)
  to retrieve performance of original shallow with the Intel
  compiler.}
\label{FIG_time_smooth_code}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=90mm]{opt_stages_256}
\caption{The performance of the \psykal version of shallow for the
  $256^{2}$ domain at each stage of optimisation. Results are given as
  a percentage of the performance of the original code for each
  compiler/CPU combination. A figure greater than 100\% indicates that
  the \psykal version performs better than the original.}
\label{FIG_opt_stages_256}
\end{figure}

>>>>

Having show that, in general, we can recover and occasionally improve
upon the performance of the original version of shallow, the next
logical step is to examine the necessary code transformations.  We do
this for the $256^{2}$ case since this fits within cache on both the
Haswell and Ivy Bridge CPUs we are using here.
Table~\ref{TABLE_opt_breakdown} shows detailed timings for this case
after each transformation has been applied to the code. The same data
is visualised in Figure~\ref{FIG_opt_stages_256} where the performance
of the \psykal version for a given compiler/CPU is given relative to
the performance of the original with the same compiler/CPU.

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Compiler:          &  Gnu    & Intel   & Gnu & Intel & Cray & IBM bgxlf \\
\hline
CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
\hline
Original                             & 1.49 & 1.41 & 1.54 & 1.57 & 1.23 &        \\
Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.96 & 1.67 &        \\
Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.93 & 1.39 & 113.74 \\
Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.93 & 1.42 &  68.28 \\
Loops fused                          & 2.18 & 1.58 & 1.79 & 1.76 & 1.47 &  71.91 \\
In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.75 & 1.46 &  83.36 \\
Fused field-copy                     & 2.18 & 1.85 & 1.99 & 2.07 & 1.30 &  72.62 \\
In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.59 & 1.30 &        \\
Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.55 & 1.25 &        \\
Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.54 & 1.23 &        \\
\hline
\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
\hline
\end{tabular}
\end{table*}

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called has the most dramatic
effect on the performance of the compiled code; it now achieves almost
80\% of the performance of the original. The next most significant
transformation is that of (manually) in-lining the kernel code itself into
the bodies of the loops from which they are called. This gets the
\psykal version to within 8\% of the performance of the original.
{\bf What's losing us 8\% of performance with the Gnu compiler?}

With the Intel compiler, the key transformations to recover
performance are different (second and fourth clusters in
Figure~\ref{FIG_opt_stages_256}).  Table~\ref{TABLE_opt_breakdown}
shows that even the vanilla \psykal version of shallow achieves some
80\% of the performance of the original version. Only two code
transformations were required to increase this to 100\%. The first of
these was the fusion of the computational loops. However, we found
that performance suffered (purple bar) when this was extended to
include the loop required for field copies at the end of a time-step
(Figure~\ref{FIG_time_smooth_code}). Further investigation of this
slow-down revealed that the compiler was unable to SIMD vectorise the
newly-fused loop after it had in-lined the body of the kernel called
from within it. Once this in-lining was done by hand (dark-blue bar),
the compiler was able to vectorise the loop and the performance of the
original version of shallow was recovered.

\begin{figure}[!t]
\centering
\includegraphics[width=95mm]{opt_stages_256}
\caption{The performance of the \psykal version of shallow for the
  $256^{2}$ domain at each stage of optimisation. Results are given as
  a percentage of the performance of the original code for each
  compiler/CPU combination. A figure greater than 100\% indicates that
  the \psykal version performs better than the original.}
\label{FIG_opt_stages_256}
\end{figure}

\begin{figure}
\begin{verbatim}
! Time smoothing
DO J=1,N+1
  DO I=1,M+1
    CALL time_smooth_code(i,j,ufld,unew,uold)
!    uold(i,j) = ufld(i,j) + alpha* &
!     (unew(i,j) - 2.*ufld(i,j) + uold(i,j))
  END DO
END DO

! Copy new fields into current fields
DO J=1,N+1
  DO I=1,M+1
    Ufld(I,J) = UNEW(I,J)
  END DO
END DO
\end{verbatim}
\caption{Example of the coding structure that required manual
  in-lining of the kernel body (as indicated by commented-out lines)
  to retrieve performance of original shallow with the Intel
  compiler.}
\label{FIG_time_smooth_code}
\end{figure}

The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
evolution of the performance of the \psykal version of shallow with
the Cray compiler. Again, this has significant differences from the
behaviour seen with the Gnu and Intel compilers. As with the Intel
compiler, the performance of the Vanilla \psykal version is fairly
good at 74\% of that of the original. In contrast to the other
compilers consisdered so far, the first significant transformation is
simply to specify the bounds of the arrays being used within the
middle layer using module variables (as opposed to specifying them as
assumed size). This tells the compiler that all of the arrays used in
the middle layer have the same extent and also that all of the
computational loops are over almost all elements of these arrays. This
simple change gives a \psykal version that achieves 88\% of the
performance of the original.

Subsequent transformations actually harm performance until the
field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
with the preceeding loop. The resulting \psykal version now achieves
94\% of the original performance. Two further steps are required to
match the latter. First, the first loop nest has to be fully fused
(both inner and outer loops fused) and second, the array/loop bounds
are passed as an argument to the middle layer. This is done in order
to specify to the compiler that these bounds remain constant for the
duration of the time-stepping loop.
